Alternating Automata Modulo First Order Theories

### 2019-02-07 quantifiers/body.tex
ok

We check emptiness of first order alternating automata using a version
of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
analogous procedure is given in \cite{IosifXu18}, for a simpler model
of alternating automata, that uses only predicates or arity zero
(propositional variables) and no transition quantifiers. For
simplicity, we defer the details of the semi-algorithm to Appendix
\ref{app:algorithm} and content ourselves of several high-level
definitions.

@@

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels of (some of the) prefixes of $\alpha$ (and
that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
indicating that there is no real execution of the automaton along that
input event sequence. As a result of the change of labels, if a node
$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
might not cover it with the new label. Therefore, the coverage
relation has to be recomputed after each refinement of the
labeling. The semi-algorithm stops when (and if) a safe complete
unfolding has been found. For a detailed presentation of the emptiness
procedure, we refer to \cite{IosifXu18}.

@@

Typically, when checking the unreachability of a set of program
configurations \cite{McMillan06}, the interpolants used to annotate
the unfolded control structure are assertions about the values of the
program variables in a given control state, at a certain step of an
execution. Because we consider alternating computation trees
(forests), we must distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are formulae
involving a single predicate symbol, whereas \emph{time-local}
interpolants involve only predicates $\stamp{q}{i}$ and variables
$\stamp{x}{i}$, for a single $i \geq 0$.

@@

In the rest of this section, let us fix an automaton $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above observation, none
of the interpolants considered will be control-local and we shall use
the term \emph{local} to denote time-local interpolants, with no free
variables. The following definition generalizes interpolants from
unsatisfiable conjunctions to sequences of input events:

@@

This is where the predicate-free path formulae (defined in
\S\ref{sec:path-formulae}) come into play. Recall that, for a given
event sequence $\alpha$, the automaton $\A$ accepts a word $w$ such
that $\event{w} = \alpha$ if and only if $\substform{\alpha}$ is
satisfiable (Lemma \ref{lemma:quant-pred-acceptance}). Assuming
further that the equality and interpreted predicates
(e.g.\ inequalities for integers) atoms from the transition rules of
$\A$ belong to a decidable first order theory, such as Presburger
arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives us an
effective way of checking emptiness of $\A$, relative to a given event
sequence. However, this method does not cope well with lazy
annotation, because there is no way to extract, from the
unsatisfiability proof of $\substform{\alpha}$, the interpolants
needed to annotate $\alpha$. This is because \begin{inparaenum}[(i)]
\item\label{it1:gli} the formula $\substform{\alpha}$, obtained by
  repeated substitutions loses track of the steps of the execution,
  and
%
\item\label{it2:gli} quantifiers that occur nested in
  $\substform{\alpha}$ make it difficult to write $\substform{\alpha}$
  as an unsatisfiable conjunction of formulae from which interpolants
  are extracted (Definition \ref{def:lyndon-interpolant}).
\end{inparaenum}

@@

\begin{theorem}[\cite{BorgerGraedelGurevich97}]\label{thm:skolem}
  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence,
  where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
  quantifier-free, let $\eta_i \isdef f_i(y_1,\ldots,y_{k_i})$ if $Q_i
  = \forall$ and $\eta_i \isdef x_i$ if $Q_i = \exists$, where $f_i$
  is a fresh function symbol and $\set{y_1, \ldots, y_{k_i}} =
  \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1 x_1
  \ldots Q_n x_n ~.~ \phi \models \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\)
  holds.
\end{theorem}
\paragraph{\em Proof.} See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed

Examples of witness-producing quantifier elimination procedures can be
found in the literature for e.g. linear integer (real) arithmetic
(LIA,LRA), Presburger arithmetic and boolean algebra of sets and
Presburger cardinality constraints (BAPA)
\cite{KuncakMayerPiskacSuter12}.

@@

Under the assumption that witness terms can be effectively built, let
us describe the generation of a non-local GLI for a given input event
sequence $\alpha = a_1 \ldots a_n$. First, we generate successively
the acceptance formula $\accform{\alpha}$ and its equisatisfiable
forms $\quantform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$
and $\substform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$,
both written in prenex form, with matrices $\widehat{\Phi}$ and
$\overline{\Phi}$, respectively. Because we assumed that the first
order theory of $\Data$ has quantifier elimination, the satisfiability
problem for $\substform{\alpha}$ is decidable. If $\substform{\alpha}$
is satisfiable, we build a counterexample for emptiness $w$ such that
$\event{w}=\alpha$ and $\data{w}$ is a satisfying assignment for
$\substform{\alpha}$. Otherwise, $\substform{\alpha}$ is unsatisfiable
and there exist witness terms $\tau_{i_1} \ldots \tau_{i_\ell}$, where
$\set{i_1, \ldots, i_\ell} = \set{j \in [1,m] \mid Q_j = \forall}$,
such that $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
  \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable (Theorem
\ref{thm:skolem}). Then it turns out that the formula
$\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
  \tau_{i_\ell}/x_{i_\ell}]$, obtained analogously from the matrix of
$\quantform{\alpha}$, is unsatisfiable as well (Lemma
\ref{lemma:subst-quant}). Because this latter formula is structured as
a conjunction of formulae $\stamp{\iota}{0} \wedge \phi_1 \ldots
\wedge \phi_n \wedge \psi$, where $\voc{\phi_k} \cap \stamp{Q}{\leq n}
\subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k}$ and $\voc{\psi} \cap
\stamp{Q}{\leq n} \subseteq \stamp{Q}{n}$, it is now possible to use
an existing interpolation procedure for the quantifier-free theory of
$\Data$, extended with uninterpreted function symbols, to compute a
(not necessarily local) GLI $(I_0, \ldots, I_n)$ such that $\voc{I_k}
\cap \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$, for all $k \in [n]$.

@@

In conclusion, under two assumptions about the first order theory of
the data domain, namely the\ \begin{inparaenum}[(i)]
\item witness-producing quantifier elimination, and
\item Lyndon interpolation for the quantifier-free fragment with
  uninterpreted functions,
\end{inparaenum}
we developped a rather generic method that produces GLIs for
unfeasible input event sequences. Moreover, each formula $I_k$ in the
interpolant refers only to the current predicate symbols
$\stamp{Q}{I_k}$, the current and past input variables $\stamp{X}{\leq
  k}$ and the existentially quantified transition variables introduced
at the previous steps $\set{x_j \mid j < \xi_{\max}^{-1}(k), Q_j =
  \exists}$. The remaining question is how to use such non-local GLIs
to label the sequences in the unfolding of an automaton (Definition
\ref{def:unfolding}) and to compute the coverage (Definition
\ref{def:coverage}) between nodes of this unfolding.

@@

\begin{lemma}\label{lemma:refinement}
  Let $U$ be an unfolding of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ such that $\alpha = a_1\ldots a_n
  \in \dom(U)$ is an input event sequence and $(I_0, \ldots, I_n)$ be
  a GLI for $\alpha$. Then the following mapping $U' : \dom(U)
  \rightarrow \posforms(Q,\emptyset)$ is an unfolding of $\A$, where:
  \begin{compactitem}
  \item $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, for all $k \in [n]$,
    where $J_k$ is the formula obtained from $I_k$ by replacing each
    time-stamped predicate symbol $\stamp{q}{k}$ by $q$ and
    existentially quantifying each free variable in $I_k$,
  %
\item $U'(\beta) = U(\beta)$ if $\beta \in \dom(U)$ and $\beta
  \not\prefix \alpha$,
  \end{compactitem}
  Moreover, $\alpha$ is safe in $U'$.
\end{lemma}
\proof{
  The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
  with Definition \ref{def:unfolding}, because: \begin{compactitem}
  \item $U'(\alpha_0) \equiv \iota$, since, by point
    \ref{it2:generalized-lyndon-interpolant} of Definition
    \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
    \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
    U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
    %
  \item by Definition \ref{def:generalized-lyndon-interpolant}
    (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
    [n-1]$:
    \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
    \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
    $\overstamp{I}{j}_k$ for the formula in which each predicate
    symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
    following entailment holds:
    \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
    Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
    existentially quantifying all the free variables of $I_k$, we also
    obtain, applying Fact \ref{fact:ex-entail} below:
    \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
    satisfies the labeling condition of Definition \ref{def:unfolding}
    (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
    we obtain, as required: 
    \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models
    \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
  \end{compactitem}

@@

Observe that, by Lemma \ref{lemma:subst-quant}
(\ref{it2:subst-quant}), the set of free variables of a GLI formula
$I_k$ consists of \begin{inparaenum}[(i)]
\item variables $\stamp{X}{\leq k}$ keeping track of data values seen
  in the input at some earlier moment in time, and
%
\item variables that track past choices made within the transition
  rules.
\end{inparaenum}
Basically, it is not important when exactly in the past a certain
input has been read or when a choice has been made, as only the
relation between the values of these and the current variables
determines the future behavior of the automaton. Intuitively,
quantifying these variables existentially does the job of ignoring
when exactly in the past these values have been seen. Moreover, the
last point of Lemma \ref{lemma:refinement} ensures that the refined
path is safe in the new unfolding and thus will stay safe in all
future unfoldings of $\A$.

@@

We have implemented a version of the IMPACT semi-algorithm
\cite{McMillan06} in a prototype tool called FOADA, which is avaliable
online \cite{foada}. The tool is written in Java and uses the Z3 SMT
solver \cite{z3}, via the JavaSMT interface \cite{javasmt}, for
spuriousness and coverage queries and also for interpolant generation.
The experiments were carried out on a MacOS x64 - 1.3 GHz Intel Core
i5 - 8 GB 1867 MHz LPDDR3 machine.

\begin{table}[htb]
%\vspace*{-\baselineskip}
\begin{center}
{\fontsize{7}{8}\selectfont
\begin{tabular}{||l|c|c|c|c|c|c|c|c||}
\hline
Example & $\len{\A}$ (bytes) & Nb of Predicates & Nb of Variables & Nb of Transitions & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Time (ms) \\
\hline
incdec.pa & 499 & 3 & 1 & 12 & no & 21 & 17 & 779 \\
\hline
localdec.pa & 678 & 4 & 1 & 16 & no & 49 & 35 & 1814 \\
\hline
ticket.pa & 4250 & 13 & 1 & 73 & no & 229 & 91 & 9543 \\
\hline
count\_thread0.pa & 9767 & 14 & 1 & 126 & no & 154 & 128 & 8553 \\
\hline
count\_thread1.pa & 10925 & 15 & 1 & 135 & no & 766 & 692 & 76771 \\
\hline
local0.pa & 10595 & 13 & 1 & 117 & no & 73 & 27 & 1431 \\
\hline
local1.pa & 11385 & 14 & 1 & 126 & no & 1135 & 858 & 101042 \\
\hline
array\_rotation.ada & 1834 & 8 & 7 & 7 & yes & 9 & 8 & 1543 \\
\hline
array\_simple.ada & 3440 & 9 & 5 & 8 & yes & 11 & 10 & 6787 \\
\hline
array\_shift.ada & 874 & 6 & 5 & 5 & yes & 6 & 5 & 413 \\
\hline
abp.ada & 6909 & 16 & 14 & 28 & no & 52 & 47 & 4788 \\
\hline
train.ada & 1823 & 10 & 4 & 26 & yes & 68 & 67 & 7319 \\
\hline
hw1.ada & 322 & 3 & 2 & 5 & Solver Error & / & / & / \\
\hline
hw2.ada & 674 & 7 & 2 & 8 & yes & 20 & 22 & 4974 \\
\hline
rr-crossing.foada & 1780 & 10 & 1 & 16 & yes & 67 & 67 & 7574 \\
% \hline
% running2.foada & 1217 & yes & 13 & 13 & 788 \\
\hline
train-simple1.foada & 5421 & 13 & 1 & 61 & yes & 43 & 44 & 2893 \\
\hline
train-simple2.foada & 10177 & 16 & 1 & 118 & yes & 111 & 113 & 8386 \\
\hline
train-simple3.foada & 15961 & 19 & 1 & 193 & yes & 196 & 200 & 15041 \\
% \hline
% fischer-mutex1.foada & 1566 & no & 31 & 31 & 1552 \\
\hline
fischer-mutex2.foada & 3000 & 11 & 2 & 23 & yes & 23 & 23 & 808 \\
\hline
fischer-mutex3.foada & 4452 & 16 & 2 & 34 & yes & 33 & 33 & 1154 \\
\hline
\end{tabular}
}
\caption{Experiments with First Order Alternating Automata}
\label{tab:experiments}
\end{center}
%\vspace*{-3\baselineskip}
\end{table}

The experimental results, reported in Table \ref{tab:experiments},
come from several sources, namely predicate automata models (*.pa)
\cite{Farzan15,Farzan16} available online \cite{pa}, timed automata
inclusion problems ({\tt abp.ada}, {\tt train.ada}, {\tt
  rr-crossing.foada}), array logic entailments ({\tt
  array\_rotation.ada}, {\tt array\_simple.ada}, {\tt
  array\_shift.ada}) and hardware circuit verification ({\tt hw1.ada},
{\tt hw2.ada}), initially considered in \cite{IosifRV16}. The {\tt
  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.foada} examples are
parametric verification problems in which one checks inclusions of the
form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$, where $A_i$ is
the $i$-th copy of the same template automaton.

>>>>

We check emptiness of first order alternating automata using a version
of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
analogous procedure is given in \cite{IosifXu18}, for a simpler model
of alternating automata, that uses only predicates or arity zero
(propositional variables) and no transition quantifiers. For
simplicity, we defer the pseudocode of the semi-algorithm to Appendix
\ref{app:algorithm} and give here several high-level definitions.

@@

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels of (some of the) prefixes of $\alpha$ (and
that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
indicating that there is no real execution of the automaton along that
input event sequence. As a result of the change of labels, if a node
$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
might not cover it with the new label. Therefore, the coverage
relation has to be recomputed after each refinement of the
labeling. The semi-algorithm stops when (and if) a safe complete
unfolding has been found. For a detailed presentation of the emptiness
procedure, we refer to Appendix \ref{app:algorithm}.

@@

Typically, when checking the unreachability of a set of program
configurations, the interpolants used to annotate the unfolded control
structure are assertions about the values of the program variables in
a given control state, at a certain step of an execution
\cite{McMillan06}. Because we consider alternating computation trees
(forests), we must distinguish between
\begin{inparaenum}[(i)]
\item locality of interpolants w.r.t. a given control state (control
  locality) and
%
\item locality w.r.t. a given time stamp (time locality).
\end{inparaenum} 
In logical terms, \emph{control-local} interpolants are formulae
involving a single predicate symbol, whereas \emph{time-local}
interpolants involve only predicates $\stamp{q}{i}$ and variables
$\stamp{x}{i}$, for a single $i \geq 0$.

@@

In the rest of this section, fix an automaton $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above remark, none of
the interpolants considered will be control-local and we shall use the
term \emph{local} to denote time-local interpolants, with no free
variables. The following definition generalizes interpolants from
unsatisfiable conjunctions to input sequences:

@@

This is where the predicate-free path formulae (defined in
\S\ref{sec:path-formulae}) come into play. Recall that, for a given
event sequence $\alpha$, the automaton $\A$ accepts a word $w$ such
that $\event{w} = \alpha$ if and only if $\substform{\alpha}$ is
satisfiable (Lemma \ref{lemma:quant-pred-acceptance}). Assuming
further that the equality and interpreted predicates
(e.g.\ inequalities for integers) atoms from the transition rules of
$\A$ belong to a decidable first order theory, such as Presburger
arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives us an
effective way of checking emptiness of $\A$, relative to a given event
sequence. However, this method does not cope well with lazy
annotation, because there is no way to extract, from the
unsatisfiability proof of $\substform{\alpha}$, the interpolants
needed to annotate $\alpha$. This is because \begin{inparaenum}[(I)]
\item\label{it1:gli} the formula $\substform{\alpha}$, obtained by
  repeated substitutions loses track of the steps of the execution,
  and
%
\item\label{it2:gli} quantifiers that occur nested in
  $\substform{\alpha}$ make it difficult to write $\substform{\alpha}$
  as an unsatisfiable quantifier-free conjunction of formulae from
  which interpolants are extracted (Definition
  \ref{def:lyndon-interpolant}).
\end{inparaenum}

@@

\begin{theorem}[\cite{BorgerGraedelGurevich97}]\label{thm:skolem}
  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence,
  where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
  quantifier-free, let $\eta_i \isdef f_i(y_1,\ldots,y_{k_i})$ if $Q_i
  = \forall$ and $\eta_i \isdef x_i$ if $Q_i = \exists$, where $f_i$
  is a fresh function symbol and $\set{y_1, \ldots, y_{k_i}} =
  \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1 x_1
  \ldots Q_n x_n ~.~ \phi \models \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\)
  holds.
\end{theorem}
\paragraph{\em Proof.} See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
\cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed
\noindent
Examples of witness-producing quantifier elimination procedures can be
found in the literature for e.g. linear integer (real) arithmetic
(LIA,LRA), Presburger arithmetic and boolean algebra of sets and
Presburger cardinality constraints (BAPA)
\cite{KuncakMayerPiskacSuter12}.

@@

Under the assumption that witness terms can be effectively built, we
describe the generation of a non-local GLI for a given input event
sequence $\alpha = a_1 \ldots a_n$. First, we generate successively
the acceptance formula $\accform{\alpha}$ and its equisatisfiable
forms $\quantform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$
and $\substform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$,
both written in prenex form, with matrices $\widehat{\Phi}$ and
$\overline{\Phi}$, respectively. Because we assumed that the first
order theory of $\Data$ has quantifier elimination, the satisfiability
problem for $\substform{\alpha}$ is decidable. If $\substform{\alpha}$
is satisfiable, we build a counterexample for emptiness $w$ such that
$\event{w}=\alpha$ and $\data{w}$ is a satisfying assignment for
$\substform{\alpha}$. Otherwise, $\substform{\alpha}$ is unsatisfiable
and there exist witness terms $\tau_{i_1} \ldots \tau_{i_\ell}$, where
$\set{i_1, \ldots, i_\ell} = \set{j \in [1,m] \mid Q_j = \forall}$,
such that $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
  \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable (Theorem
\ref{thm:skolem}). Then it turns out that the formula
$\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
  \tau_{i_\ell}/x_{i_\ell}]$, obtained analogously from the matrix of
$\quantform{\alpha}$, is unsatisfiable as well (Lemma
\ref{lemma:subst-quant} below). Because this latter formula is
structured as a conjunction of formulae $\stamp{\iota}{0} \wedge
\phi_1 \ldots \wedge \phi_n \wedge \psi$, where $\voc{\phi_k} \cap
\stamp{Q}{\leq n} \subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k}$ and
$\voc{\psi} \cap \stamp{Q}{\leq n} \subseteq \stamp{Q}{n}$, it is now
possible to use an existing interpolation procedure for the
quantifier-free theory of $\Data$, extended with uninterpreted
function symbols, to compute a (not necessarily local) GLI $(I_0,
\ldots, I_n)$ such that $\voc{I_k} \cap \stamp{Q}{\leq n} \subseteq
\stamp{Q}{k}$, for all $k \in [n]$.

@@

Consequently, under two assumptions about the first order theory of
the data domain, namely\ \begin{inparaenum}[(i)]
\item witness-producing quantifier elimination, and
\item Lyndon interpolation for the quantifier-free fragment with
  uninterpreted functions,
\end{inparaenum}
we developped a generic method that produces GLIs for unfeasible input
event sequences. Moreover, each formula $I_k$ in the interpolant
refers only to the current predicate symbols $\stamp{Q}{I_k}$, the
current and past input variables $\stamp{X}{\leq k}$ and the
existentially quantified transition variables introduced at the
previous steps $\set{x_j \mid j < \xi_{\max}^{-1}(k), Q_j =
  \exists}$. The remaining questions are how to use these GLIs to
label the sequences in the unfolding of an automaton (Definition
\ref{def:unfolding}) and compute coverage (Definition
\ref{def:coverage}) between nodes of the unfolding.

@@

\begin{lemma}\label{lemma:refinement}
  Let $U$ be an unfolding of an automaton $\A =
  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ such that $\alpha = a_1\ldots a_n
  \in \dom(U)$ and $(I_0, \ldots, I_n)$ is a GLI for $\alpha$. Then
  the mapping $U' : \dom(U) \rightarrow \posforms(Q,\emptyset)$ is an
  unfolding of $\A$, where:
  \begin{compactitem}
  \item $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, for all $k \in [n]$,
    where $J_k$ is the formula obtained from $I_k$ by removing the
    time stamp of each predicate symbol $\stamp{q}{k}$ and
    existentially quantifying each free variable in $I, _k$,
  %
\item $U'(\beta) = U(\beta)$ if $\beta \in \dom(U)$ and $\beta
  \not\prefix \alpha$,
  \end{compactitem}
  Moreover, $\alpha$ is safe in $U'$.
\end{lemma}
\proof{
  The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
  with Definition \ref{def:unfolding}, because: \begin{compactitem}
  \item $U'(\alpha_0) \equiv \iota$, since, by point
    \ref{it2:generalized-lyndon-interpolant} of Definition
    \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
    \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
    U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
    %
  \item by Definition \ref{def:generalized-lyndon-interpolant}
    (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
    [n-1]$:
    \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
    \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
    $\overstamp{I}{j}_k$ for the formula in which each predicate
    symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
    following entailment holds:
    \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
    Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
    existentially quantifying all the free variables of $I_k$, we also
    obtain, applying Fact \ref{fact:ex-entail} below:
    \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
    satisfies the labeling condition of Definition \ref{def:unfolding}
    (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
    we obtain, as required: 
    \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
    \rightarrow \stamp{\psi}{1} \models
    \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
  \end{compactitem}

@@

Observe that, by Lemma \ref{lemma:subst-quant}
(\ref{it2:subst-quant}), the set of free variables of a GLI formula
$I_k$ consists of \begin{inparaenum}[(i)]
\item variables $\stamp{X}{\leq k}$ keeping track of data values seen
  in the input at some earlier moment in time, and
%
\item variables that track past choices made within the transition
  rules.
\end{inparaenum}
Basically, it is not important when exactly in the past a certain
input has been read or when a choice has been made, because only the
relation between the values of these and the current variables
determines the future behavior of the automaton. Quantifying these
variables existentially does the job of ignoring when exactly in the
past these values have been seen. Moreover, the last point of Lemma
\ref{lemma:refinement} ensures that the refined path is safe in the
new unfolding and will stay safe in all future refinements of this
unfolding.

@@

We have implemented a version of the IMPACT semi-algorithm
\cite{McMillan06} in a prototype tool, avaliable
online\footnote{\url{http://nts.imag.fr/images/e/ea/Cav19b.tar.gz}}. The
tool is written in Java and uses the Z3 SMT solver \cite{z3}, via the
JavaSMT interface \cite{javasmt}, for spuriousness and coverage
queries and also for interpolant generation. Table
\ref{tab:experiments} reports the size of the input automaton in
bytes, the numbers of Predicates, Variables and Transitions, the
result of emptiness check, the number of Expanded and Visited Nodes
during the unfolding and the Time in miliseconds.  The experiments
were carried out on a MacOS x64 - 1.3 GHz Intel Core i5 - 8 GB 1867
MHz LPDDR3 machine.

\begin{table}[htb]
\vspace*{-\baselineskip}
\begin{center}
{\fontsize{7}{8}\selectfont
\begin{tabular}{||l|c|c|c|c|c|c|c|c||}
\hline
Example & $\len{\A}$ (bytes) & Predicates & Variables & Transitions & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Time (msec) \\
\hline
incdec.pa & 499 & 3 & 1 & 12 & no & 21 & 17 & 779 \\
\hline
localdec.pa & 678 & 4 & 1 & 16 & no & 49 & 35 & 1814 \\
\hline
ticket.pa & 4250 & 13 & 1 & 73 & no & 229 & 91 & 9543 \\
\hline
count\_thread0.pa & 9767 & 14 & 1 & 126 & no & 154 & 128 & 8553 \\
\hline
count\_thread1.pa & 10925 & 15 & 1 & 135 & no & 766 & 692 & 76771 \\
\hline
local0.pa & 10595 & 13 & 1 & 117 & no & 73 & 27 & 1431 \\
\hline
local1.pa & 11385 & 14 & 1 & 126 & no & 1135 & 858 & 101042 \\
\hline
array\_rotation.ada & 1834 & 8 & 7 & 7 & yes & 9 & 8 & 1543 \\
\hline
array\_simple.ada & 3440 & 9 & 5 & 8 & yes & 11 & 10 & 6787 \\
\hline
array\_shift.ada & 874 & 6 & 5 & 5 & yes & 6 & 5 & 413 \\
\hline
abp.ada & 6909 & 16 & 14 & 28 & no & 52 & 47 & 4788 \\
\hline
train.ada & 1823 & 10 & 4 & 26 & yes & 68 & 67 & 7319 \\
\hline
hw1.ada & 322 & 3 & 2 & 5 & Solver Error & / & / & / \\
\hline
hw2.ada & 674 & 7 & 2 & 8 & yes & 20 & 22 & 4974 \\
\hline
rr-crossing.foada & 1780 & 10 & 1 & 16 & yes & 67 & 67 & 7574 \\
% \hline
% running2.foada & 1217 & yes & 13 & 13 & 788 \\
\hline
train-simple1.foada & 5421 & 13 & 1 & 61 & yes & 43 & 44 & 2893 \\
\hline
train-simple2.foada & 10177 & 16 & 1 & 118 & yes & 111 & 113 & 8386 \\
\hline
train-simple3.foada & 15961 & 19 & 1 & 193 & yes & 196 & 200 & 15041 \\
% \hline
% fischer-mutex1.foada & 1566 & no & 31 & 31 & 1552 \\
\hline
fischer-mutex2.foada & 3000 & 11 & 2 & 23 & yes & 23 & 23 & 808 \\
\hline
fischer-mutex3.foada & 4452 & 16 & 2 & 34 & yes & 33 & 33 & 1154 \\
\hline
\end{tabular}
}
\caption{Experiments with First Order Alternating Automata}
\label{tab:experiments}
\end{center}
\vspace*{-3\baselineskip}
\end{table}

The test cases shown in Table \ref{tab:experiments}, come from several
sources, namely predicate automata models (*.pa)
\cite{Farzan15,Farzan16} available online \cite{pa}, timed automata
inclusion problems ({\tt abp.ada}, {\tt train.ada}, {\tt
  rr-crossing.foada}), array logic entailments ({\tt
  array\_rota-} {\tt tion.ada}, {\tt array\_simple.ada}, {\tt
  array\_shift.ada}) and hardware circuit verification ({\tt hw1.ada},
{\tt hw2.ada}), initially considered in \cite{IosifRV16}. The {\tt
  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.foada} examples
are parametric verification problems in which one checks inclusions of
the form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$, where $A_i$
is the $i$-th copy of the same template automaton.
